---
title: "Exploratory and Descriptive Statistics and Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{exploratory-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To start, load the package.

```{r setup}
library(JWileymisc)
library(ggplot2)
library(data.table)
```

# Descriptive Statistics

The `egltable()` function calculates basic descriptive statistics.

```{r, eval = FALSE, echo = TRUE, results = "hide"}

egltable(c("mpg", "hp", "qsec", "wt", "vs"),
         data = mtcars)

``` 


```{r, echo = FALSE, results = "asis"}

pander::pandoc.table(
          egltable(c("mpg", "hp", "qsec", "wt", "vs"),
                   data = mtcars),
          caption = "Example descriptive statistics table.",
          justify = "left")

```


The `strict` argument can be used if variables are categorical but are
not coded as factors. In this case, `vs` has two levels: 0 and 1 and
the frequency and percentage of each are shown instead of the mean and
standard deviation.

```{r, eval = FALSE, echo = TRUE, results = "hide"}

egltable(c("mpg", "hp", "qsec", "wt", "vs"),
         data = mtcars, strict=FALSE)

```

```{r, echo = FALSE, results = "asis"}

pander::pandoc.table(
          egltable(c("mpg", "hp", "qsec", "wt", "vs"),
                   data = mtcars, strict=FALSE),
          caption = "Example descriptive statistics table with automatic categorical variables.",
          justify = "left")

```


`egltable()` also allows descriptive statistics to be broken down by
another variable by using the `g` argument. This not only separates
results by group but also calculates bivariate tests of the
differences between groups and effect sizes. For example, t-tests for
continuous variables and two groups or chi-square tests for
categorical variables. For more than two groups, ANOVAs are used. 

```{r, eval = FALSE, echo = TRUE, results = "hide"}

egltable(c("mpg", "hp", "qsec", "wt", "vs"), 
  g = "am", data = mtcars, strict = FALSE)

```


```{r, echo = FALSE, results = "asis"}

pander::pandoc.table(
          egltable(c("mpg", "hp", "qsec", "wt", "vs"), 
                   g = "am", data = mtcars, strict = FALSE),
          caption = "Example descriptive statistics table by group.",
          justify = "left")

```

For very skewed continuous variables, non-parametric statistics and
tests may be more appropriate. These can be generated using the
`parametric` argument. For chi-square tests with small cell sizes,
simulated p-values also can be generated.

```{r, eval = FALSE, echo = TRUE, results = "hide"}

egltable(c("mpg", "hp", "qsec", "wt", "vs"), 
         g = "am", data = mtcars, strict = FALSE,
         parametric = FALSE)

```

```{r, echo = FALSE, results = "asis"}

pander::pandoc.table(
          egltable(c("mpg", "hp", "qsec", "wt", "vs"), 
                   g = "am", data = mtcars, strict = FALSE,
                   parametric = FALSE),
          caption = "Example descriptive statistics table by group.",
          justify = "left")

```


## Paired Data

We have already seen how to compare descriptives across groups when
the groups were independent. `egltable()` also supports using groups
to test paired samples. To use this, the variable passed to the
grouping argument, `g` must have exactly two levels and you must also
pass a variable that is a unique ID per unit and specify `paired =
 TRUE`.


By default for continuous, paired data, mean and standard deviations
are presented and a paired samples t-test is used. A pseudo Cohen's d
effect size is calculated as the mean of the change score divided by
the standard deviation of the change score. If there are missing data,
its possible that the mean difference will be different than the
difference in means as the means are calculated on all available data,
but the effect size can only be calculated on complete cases.

```{r, eval = FALSE, echo = TRUE, results = "hide"}
## example with paired data
egltable(
  vars = "extra",
  g = "group",
  data = sleep,
  idvar = "ID",
  paired = TRUE)

```

```{r, echo = FALSE, results = "asis"}

pander::pandoc.table(
egltable(
  vars = "extra",
  g = "group",
  data = sleep,
  idvar = "ID",
  paired = TRUE),
caption = "Example parametric descriptive statistics for paired data.",
justify = "left")

```

If we do not want to make parametric assumptions with continuous
variables, we can set `parametric = FALSE`. In this case the
descriptives are medians and a paired Wilcoxon test is used. In this
dataset there are ties and a warning is generated about ties and
zeroes. This warning is generally ignorable, but if these were central 
hypothesis tests, it may warrant further testing using, for example,
simulations which are more precise in the case of ties.

```{r, eval = FALSE, echo = TRUE, results = "hide"}
egltable(
  vars = "extra",
  g = "group",
  data = sleep,
  idvar = "ID",
  paired = TRUE,
  parametric = FALSE)

```

```{r, echo = FALSE, results = "asis"}

pander::pandoc.table(
egltable(
  vars = "extra",
  g = "group",
  data = sleep,
  idvar = "ID",
  paired = TRUE,
  parametric = FALSE),
caption = "Example non parametric descriptive statistics for paired data.",
justify = "left")

```

We can also work with categorical paired data. The following code
creates a categorical variable, the tertiles of chick weights measured
over time. The chick weight dataset has many time points, but we will
just use two.

```{r}

## paired categorical data example
## using data on chick weights to create categorical data
tmp <- subset(ChickWeight, Time %in% c(0, 20))
tmp$WeightTertile <- cut(tmp$weight,
  breaks = quantile(tmp$weight, c(0, 1/3, 2/3, 1), na.rm = TRUE),
  include.lowest = TRUE)

```

No special code is needed to work with categorical
variables. `egltable()` recognises categorical variables and uses
McNemar's test, which is a chi-square of the off diagonals, which
tests whether people (or chicks in this case) change groups equally
over time or preferentially move one direction. In this case, a
significant result suggests that over time chicks' weights change
preferentially one way and the descriptive statistics show us that
there is an increase in weight tertile from time 0 to time 20.

```{r, eval = FALSE, echo = TRUE, results = "hide"}
egltable(c("weight", "WeightTertile"), g = "Time",
  data = tmp,
  idvar = "Chick", paired = TRUE)
```

```{r, echo = FALSE, results = "asis"}

pander::pandoc.table(
egltable(c("weight", "WeightTertile"), g = "Time",
  data = tmp,
  idvar = "Chick", paired = TRUE),
caption = "Continuous and categorical paired data.",
justify = "left")

```


# Correlation Summaries

For continuous variables, correlation matrices are commonly
examined. This is especially true for structural equation models or
path analyses.

The `SEMSummary()` function provides a simple way to generate these
under various options. There is a formula interface, similar to `lm()`
or other regression models. Missing data can be handled using listwise
deletion, pairwise present data, or full information maximum
likelihood (FIML). When assumptions are met, FIML is less biased and
uses all available data, and is the default.

```{r}

m <- SEMSummary(~ mpg + hp + qsec + wt, data = mtcars)

corTab <- APAStyler(m, type = "cor", stars = TRUE)

```

These correlations can be nicely formatted into a table.

```{r, echo = FALSE, results = "asis"}

pander::pandoc.table(
          corTab$table,
          caption = "Example correlation table.",
          justify = "left")

```


Plot methods exist for `SEMSummary()` objects. 
By default, above the diagonal are correlations and below the diagonal
are p-values. However, the type argument can be set (see `?corplot`)
to get all values to be either correlations or p-values. 
By default, another useful feature is that
hierarchical clustering is used to group similar variables together in
clusters, provided a more useful sorting of the data than many
"default" correlation matrices. If a specific order is desired, you
can use the `order = "asis"` option to keep the variable order the
same as written in `SEMSummary()`. 

```{r}

plot(m) +
  ggtitle("Order by hierarchical clustering")

plot(m, order = "asis") +
  ggtitle("Order as written")

``` 

```{r}

plot(m, type = "p") +
  ggtitle("Numbers are p-values")

```

## Grouped Correlations

Correlations also can be broken down by group. 
Here results are separated by species which are automaticaly 
used as the title of each graph.

```{r}

mg <- SEMSummary(~ Sepal.Length + Petal.Length +
                  Sepal.Width + Petal.Width | Species,
                 data = iris)

plot(mg)

```

# Likert Scale Plots

In much of psychological and consumer/market research, likert rating
scales are used. For example, rating a question/item from "Strongly
DISagree" to "Strongly Agree" or rating satisfaction from "Not at all"
to "Very Satisfied" or adjectives that capture mood/affect from "Not
at all" to "Extremely". Likert plots aim to show these results clearly
and aid interpretation by presenting the anchors as well.

The following code creates some simulated data, summarizes it, adds
the necessary labels/anchors, and creates a nice plot.

```{r}

## simulate some likert style data
set.seed(1234)
d <- data.table(
  Happy = sample(1:5, 200, TRUE, c(.1, .2, .4, .2, .1)),
  Cheerful = sample(1:5, 200, TRUE, c(.1, .2, .2, .4, .1)),
  Peaceful = sample(1:5, 200, TRUE, c(.1, .1, .2, .4, .2)),
  Sad = sample(1:5, 200, TRUE, c(.1, .3, .3, .2, .1)),
  Hopeless = sample(1:5, 200, TRUE, c(.3, .3, .2, .2, 0)),  
  Angry = sample(1:5, 200, TRUE, c(.4, .3, .2, .08, .02)))

dmeans <- melt(d, measure.vars = names(d))[,
  .(Mean = mean(value, na.rm = TRUE)), by = variable]

dmeans[, Low := paste0(variable, "\nNot at all")]
dmeans[, High := paste0(variable, "\nExtremely")]
dmeans[, variable := as.integer(factor(variable))]

## view the summarised data
print(dmeans)

gglikert("Mean", "variable", "Low", "High", data = dmeans,
         xlim = c(1, 5),
         title = "Average Affect Ratings")

```


```{r}

## create a grouping variable
dg <- cbind(d, Group = ifelse(
                 d$Happy > mean(d$Happy, na.rm = TRUE),
                 "General Population", "Depressed"))

dgmeans <- melt(dg, measure.vars = names(d), id.vars = "Group")[,
  .(Mean = mean(value, na.rm = TRUE)), by = .(variable, Group)]

dgmeans[, Low := paste0(variable, "\nNot at all")]
dgmeans[, High := paste0(variable, "\nExtremely")]
dgmeans[, variable := as.integer(factor(variable))]

## view the summarised data
print(dgmeans)

gglikert("Mean", "variable", "Low", "High",
         colour = "Group",
         data = dgmeans,
         xlim = c(1, 5),
         title = "Average Affect Ratings") +
  scale_colour_manual(
    values = c("Depressed" = "black",
               "General Population" = "grey70"))

``` 
