---
title: "Model Performance and Tests"
author: "Joshua F. Wiley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Performance and Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To start, load the package.

```{r setup}
library(JWileymisc)
```

# Model Performance

`modelPerformance()` is a generic function that can be used to 
calculate performance metrics for a model. `JWileymisc` implements 
methods for `lm` class objects. The output is a named list, 
with a data table containing results. For linear models, current
performance metrics include:

* AIC (Akaike Information Criterion)
* BIC (Bayesian Information Criterion)
* LL (Log Likelihood)
* LLDF (degrees of freedom for log likelihood)
* Sigma (residual standard deviation)
* R2 ($R^2$ variance accounted for in the sample)
* F2 (Cohen's $f^2$ effect size, calculated as 
  $\frac{R^{2}}{1 - R^{2}}$)
* AdjR2 (Sample size adjusted $R^2$, a better estimate of population
  variance accounted for)
* F (model F test)
* FNumDF (numerator degrees of freedom for model F test)
* FDenDF (denominator degrees of freedom for model F test)
* P (p value for model F test)

```{r}

mtcars$cyl <- factor(mtcars$cyl)
m <- stats::lm(mpg ~ hp + cyl, data = mtcars)

mp <- modelPerformance(m)
print(mp)

``` 

If only certain metrics are desired, these can be found by extracting
the "Performance" list element and then the correct column from the
data table.

```{r}

## Cohen's f^2 effect size
mp$Performance[, F2]

``` 

Another function, `modelTest()` is a generic providing a 
comprehensive series of tests for a model. Currently methods are
implemented for both `lm` class models and `vglm` class models from
the `VGAM` package with a multinomial family.

`modelTest()` 

```{r}

mt <- modelTest(m)
print(mt)

``` 


```{r}

APAStyler(mt)

``` 


The model tests can also be used with interactions.

```{r}

m2 <- stats::lm(mpg ~ hp * cyl, data = mtcars)

APAStyler(modelTest(m2))

``` 
